---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# rplaza

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/almeidasilvaf/rplaza)](https://github.com/almeidasilvaf/rplaza/issues)
<!-- badges: end -->

The goal of `rplaza` is to provide users with an R interface to the PLAZA
database of plant comparative genomics. `rplaza` can be used to retrieve PLAZA 
data (e.g., genome and locus sequences, genome annotation, functional 
annotation, homology, etc) directly from an R session using 
standard R/Bioconductor data classes, as summarized below:

```{r echo = FALSE}
summary_df <- data.frame(
    Data = c("Sequences", "Ranges", "Functions and homology"),
    Format = c("FASTA", "GFF3", "TSV"),
    `R class` = c("*AAStringSet* / *DNAStringSet*", "*GRanges*", "*data.frame*")
)

knitr::kable(summary_df, col.names = c("Data", "File format", "R/BioC class"))
```


## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). 
Then install `rplaza` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("rplaza")
```

And the development version from [GitHub](https://github.com/almeidasilvaf/rplaza) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("almeidasilvaf/rplaza")
```

## Citation

Below is the citation output from using `citation('rplaza')` in R. Please
run this yourself to check for any updates on how to cite __rplaza__.

```{r 'citation', eval = requireNamespace('rplaza')}
print(citation('rplaza'), bibtex = TRUE)
```

Please note that the `rplaza` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `rplaza` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

