---
title: "Accessing data from the PLAZA database"
author: 
  - name: Fabricio Almeida-Silva
    affiliation: VIB-UGent Center for Plant Systems Biology, Ghent, Belgium
  - name: Yves Van de Peer
    affiliation: VIB-UGent Center for Plant Systems Biology, Ghent, Belgium
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: yes
bibliography: vignette_bibliography.bib
vignette: >
  %\VignetteIndexEntry{Accessing data from the PLAZA database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL
)
```

# Introduction

PLAZA [@van2022plaza] is a database for plant comparative and functional 
genomics, centralizing data produced by different genome sequencing 
initiatives. Like Ensembl [@dyer2025ensembl], PLAZA organizes data in instances
based on clades of the (green) tree of life (PLAZA Dicots, PLAZA Monocots, 
PLAZA Diatoms, and pico-PLAZA). Besides its rich web server where users 
can interactively search and explore plant genomic data, PLAZA offers 
bulk download of data, including whole-genome and locus sequences, 
gene annotations, functional annotations, homology, etc. 
`r BiocStyle::Githubpkg("rplaza")` provides an R interface to the PLAZA 
database, allowing the direct import of PLAZA data to the R session using
standard R/Bioconductor classes, allowing seamless interoperability and 
integration with Bioconductor packages.

Data types provided by PLAZA and how they are stored for interoperability 
with R/Bioconductor packages are summarized below:

```{r echo = FALSE}
summary_df <- data.frame(
    Data = c("Sequences", "Ranges", "Functions and homology"),
    Format = c("FASTA", "GFF3", "TSV"),
    `R class` = c("*AAStringSet*/*DNAStringSet*", "*GRanges*", "*data.frame*")
)

knitr::kable(summary_df, col.names = c("Data", "File format", "R class"))
```

# Installation

`r BiocStyle::Githubpkg("rplaza")` can be installed from Bioconductor
with the following code:

```{r installation, eval=FALSE}
if(!requireNamespace('BiocManager', quietly = TRUE))
  install.packages('BiocManager')

BiocManager::install("rplaza")
```

```{r load_package, message=FALSE}
# Load package after installation
library(rplaza)

set.seed(123) # for reproducibility
```

# PLAZA instances and species (or *fantastic plants and where to find them*)

The PLAZA database is divided in four instances that are specialized in 
different clades:

1. **PLAZA Dicots:** focus on eudicot plants (*N* = 98 species);
2. **PLAZA Monocots:** focus on monocot plants (*N* = 52 species);
3. **pico-PLAZA:** focus on eukaryotic micro-organisms (*N* = 39 species);
4. **PLAZA Diatoms:** focus on diatoms (*N* = 26 species).

To access data on what species can be found in which instance, 
as well as associated metadata (taxonomic information, data sources, etc), 
you can inspect the object `plaza_metadata`, which comes 
with `r BiocStyle::Githubpkg("rplaza")`.
This object is a list of data frames, one for each PLAZA instance.

```{r}
# Load PLAZA metadata
data("plaza_metadata")

names(plaza_metadata)
```

As you can see, the list has four data frames.
If you want to have a look at the species in PLAZA Dicots, for instance, you
need to extract the corresponding data frame from the list.

```{r}
# Look at metadata for 
head(plaza_metadata$Dicots)
```

Note that PLAZA gives unique IDs to each species, and these are available
in column `PLAZA_ID` of each data frame. These PLAZA species IDs are essential
to work with `r BiocStyle::Githubpkg("rplaza")`, as most functions in this 
package require specifying species using PLAZA IDs.

It is also important to keep in mind that some species are present in multiple
PLAZA instances (e.g., *Arabidopsis thaliana*), so if you're combining these
data frames in a single data frame, you need to remove duplicate rows. 
For example:

```{r}
# Combine all data frames together and remove redundancy
plaza_all <- dplyr::bind_rows(plaza_metadata, .id = "instance") |>
    dplyr::distinct(PLAZA_ID, .keep_all = TRUE)

head(plaza_all)
nrow(plaza_all)
```

After combining all data frames together and removing duplicates, we can see
that there are `r nrow(plaza_all)` unique species across all PLAZA instances.

# A note on software design

`r BiocStyle::Githubpkg("rplaza")` uses `r BiocStyle::Biocpkg("BiocFileCache")`
to create a **temporary local cache** of remote resources per R session.
This way, if you try to retrieve the same data multiple times in a single
R session, you will only download data once. Caching is especially important
in two common scenarios:

1. You downloaded data for species A and B, but later decide to include species
C and D as well. If you re-run the code to obtain data for species A, B, C, 
and D, you will only download data for species C and D, as data for species A
and B are already stored in a local cache.

2. You tried to download data for several species, but after some time your
download failed for some reason (internet issues, timeout, etc). If you re-run
the code to download data, your download will start from where it failed, so
it will finish much faster.

# Obtaining sequence data

PLAZA provides sequences for whole genomes and whole-genome loci (all proteins,
all transcripts, and all CDS in a genome). These are available as FASTA
files, which `r BiocStyle::Githubpkg("rplaza")` reads as `DNAStringSet` objects
(or `AAStringSet` objects for proteins).

## Genomes

To extract whole-genome sequences for a particular species, you will use 
the function `get_genome()`. For a single species, `get_genome()` takes a 
species ID and returns a `DNAStringSet` object.
To demonstrate it, let's obtain the whole genome of *Arabidopsis thaliana*.

```{r}
# Retrieve genome of Arabidopsis thaliana
ath_genome <- get_genome("ath")

ath_genome
```

For multiple species, `get_genome()` takes a vector of species IDs 
and returns a list of `DNAStringSet` objects.

```{r}
# Retrieve genome of A. thaliana and A. lyrata
arabidopsis_genomes <- get_genome(c("ath", "aly"))

arabidopsis_genomes
```

To learn more about how to work with `DNAStringSet` objects,
read the vignette of the `r BiocStyle::Biocpkg("Biostrings")` package.

## Loci

To retrieve sequences for all proteins, transcripts, or CDS in the genome 
of a particular species, you will use the function `get_sequences()`. 
Besides specifying one
or multiple species IDs, `get_sequences()` also requires the locus `type` 
(one of 'CDS', 'transcript', or 'protein'), and what `transcripts` to extract
(one of 'all' or 'longest'). 

For example, researchers in evolutionary genomics often want to obtain 
protein sequences of the longest transcripts of each gene, so that they can 
compare genomes. These are usually called 'proteomes' by software tools such as
OrthoFinder [@emms2019orthofinder] and syntenet [@almeida2023syntenet]. 
To do that, you'd run:

```{r}
# Retrieve protein sequences of the longest transcripts (one per gene)
proteome <- get_sequences("ath", type = "protein", transcripts = "longest")

proteome
```

Alternatively, one might be interested in all transcripts produced by a 
particular genome. In transcriptomics, this is often called a 'reference
transcriptome', and it is required by software tools that quantify
transcript abundances, such as salmon [@patro2017salmon]. 
To do that, you would run:

```{r}
# Obtain all transcripts produced by a genome (aka 'reference transcriptome')
ref_tx <- get_sequences("ath", type = "transcript", transcripts = "all")

ref_tx
```

Note that the number of sequences in `ref_tx` is much larger than the number
of sequences in `proteome`, since the former contains sequences for all
transcripts (including multiple isoforms per gene).

To learn more about how to work with `AAStringSet` and `DNAStringSet` objects,
read the vignette of the `r BiocStyle::Biocpkg("Biostrings")` package.

# Obtaining gene annotations (a.k.a. 'ranges')

The obtain the genomic locations (ranges, with start and end coordinates) 
of all genes in the genome of a particular species, you will use the
function `get_annotation()`. This function returns a `GRanges` object
(or `GRangesList` for multiple species) with genomic coordinates
and associated metadata.
As input, you need to specify the PLAZA species
ID of your species of interest (one or multiple), what `transcripts` you
want to extract ('all' or 'longest', exactly as in `get_sequences()`),
and what `features` to extract per transcript ('all' - including exons, CDS,
UTRs, etc. - or 'exons').

For example, if you want to extract ranges for the complete genome annotation
(i.e., all transcripts, all features), you'd run:

```{r}
# Extract ranges for all features (exons, CDS, UTRs, etc) of all transcripts
annotation_all <- get_annotation("ath", transcripts = "all", features = "all")

annotation_all
```

To learn more about how to work with `GRanges` objects,
read the vignette of the `r BiocStyle::Biocpkg("GenomicRanges")` package.

# Obtaining functional annotations

PLAZA offers functional annotation for all genes in all species using
[GO terms](http://www.geneontology.org/), 
[MapMan](https://mapman.gabipd.org/) pathways, 
and [InterPro](https://www.ebi.ac.uk/interpro/) domains.
To retrieve functional annotation for all genes of a particular species, 
you will use the function `get_functional_annotation()`. As input,
this function takes one or multiple PLAZA species IDS and the `ontology`
to use (one of 'GO', 'MapMan', or 'InterPro'), and it returns a data frame
(or list of data frames, in case of multiple species) with terms associated
with each gene.

For example, to obtain InterPro domains annotated to each gene 
of *Arabidopsis thaliana*, you'd use:

```{r}
# Get InterPro domains
ath_interpro <- get_functional_annotation("ath", ontology = "InterPro")

head(ath_interpro)
```

As another example, to obtain GO annotations for the Chlorophyta algae 
*Chlamydomonas reinhardtii* and *Micromonas commoda*, you'd use:

```{r}
# Obtain functional annotation for Chlorophyta algae
algae_go <- get_functional_annotation(c("cre", "mco"), ontology = "GO")

names(algae_go)
head(algae_go$cre)
```

Lastly, to obtain MapMan annotations for all *A. thaliana* genes, you'd use:

```{r}
# Retrieve MapMan annotations for all *A. thaliana* genes
ath_mapman <- get_functional_annotation("ath", ontology = "MapMan")

head(ath_mapman)
```

# Obtaining gene IDs and descriptions

Researchers often have transcript IDs for a particular genome and want
to obtain their corresponding gene IDs. PLAZA provides correspondence tables of
transcript-to-gene ID mappings for all species, which can be very helpful
in this context. These can be retrieved with `get_tx2gene`, which takes
as input PLAZA species IDs and what `transcripts` to retrieve (one of 'all'
or 'longest').

For example, to obtain transcript-to-gene ID mappings 
for *Arabidopsis thaliana*, you'd run:

```{r}
# Get tx2gene mappings for *A. thaliana*
ath_tx2gene <- get_tx2gene("ath", transcripts = "all")

head(ath_tx2gene)
```

A similar problem occurs when researchers have gene symbols (e.g., *FLC* and
*MAF2* in *A. thaliana*) and want to extract their gene IDs. To help in this
case, PLAZA offers ID conversion tables, which can be accessed with the
function `get_id_conversions()`, as demonstrated below:

```{r}
# Get ID conversion table for *A. thaliana*
ath_ids <- get_id_conversions("ath")

head(ath_ids)
```

Lastly, researchers are often interested in short (usually one sentence) 
descriptions for genes to help them interpret their results. 
These can be obtained from PLAZA with the function `get_descriptions()`, which
takes one or multiple PLAZA species IDs as input.

For example, to obtain gene descriptions for all *A. thaliana* genes, you'd
run:

```{r}
# Get gene descriptions for *A. thaliana*
ath_descriptions <- get_descriptions("ath")

head(ath_descriptions)
```

# Obtaining homology data

For each PLAZA instance, all (coding) genes in all species are grouped in
homologous and orthologous gene families. These gene families, sometimes 
referred to as 'orthogroups' or 'hierarchical orthologous groups (HOGs)',
are the basis for comparative analyses in evolutionary genomics. 
To obtain gene family assignments for all genes in a PLAZA instance, you will
use the function `get_family_assignments()`. As input, you need to pass
the PLAZA `instance` (one of 'Dicots', 'Monocots', 'Diatoms', or 'Pico'), and
the gene family `type` (one of 'homo' for homologous, or 'ortho' for 
orthologous).

For example, to retrieve all genes and their homologous family assignments
in PLAZA Dicots, you'd run:

```{r}
# Get family assignments from PLAZA Dicots
dicots_fams <- get_family_assignments(instance = "Dicots", type = "homo")

head(dicots_fams)
```

Once you have some homologous or orthologous gene families of 
interest (e.g., families that expanded in a particular clade), you would
probably want to know the functions of the genes in these families.
Luckily for you, PLAZA stores data on overrepresented functions in its gene
families. These can be accessed with the function `get_family_functions()`, but
they are only available for PLAZA Dicots and PLAZA Monocots.

For example, to retrieve overrepresented functions in all homologous 
gene families in PLAZA Dicots, you'd run:

```{r}
# Get overrepresented functions in PLAZA Dicots families
dicots_fams_func <- get_family_functions(instance = "Dicots", type = "homo")

head(dicots_fams_func)
```

Lastly, PLAZA also stores data on block duplicates (anchor pairs in syntenic 
regions, derived from segmental or whole-genome duplications) for all its 
genomes. These data can be retrieved with 
the function `get_block_duplicates()`, which takes one or multiple PLAZA
species IDs as input.

For example, to retrieve all *A. thaliana* block duplicates, you'd run:

```{r}
# Get block duplicates for *A. thaliana*
ath_blocks <- get_block_duplicates("ath")

head(ath_blocks)
```

Note that the `Ks` column is empty, meaning that synonymous substitution rates
($K_s$) were not calculated for this genome.

# Session information {.unnumbered}

This document was created under the following conditions:

```{r, echo = FALSE}
sessioninfo::session_info()
```

# References {.unnumbered}
