---
title: "Data acquisition"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

```{r}
library(tidyverse)
library(taxize)

set.seed(123)
```

# `data/`

Here we will describe the code used to create package data in `data/`.
We will start by defining some helper functions.

```{r}
#' Format a `classification` object (from taxize) into a data frame
format_classification <- function(class_list) {
    
    class_df <- Reduce(rbind, lapply(seq_along(class_list), function(x) {
        
        cdf <- class_list[[x]]
        fn <- function(x) { return(ifelse(length(x) == 0, NA, x)) } 
        
        df <- data.frame(
            ncbi_species = fn(cdf$name[cdf$rank == "species"]),
            family = fn(cdf$name[cdf$rank == "family"]),
            order = fn(cdf$name[cdf$rank == "order"]),
            class = fn(cdf$name[cdf$rank == "class"]),
            superclass = fn(cdf$name[cdf$rank == "superclass"]),
            subphylum = fn(cdf$name[cdf$rank == "subphylum"]),
            phylum = fn(cdf$name[cdf$rank == "phylum"])
        )
        
        return(df)
    }))
    
    return(class_df)
}
```

## *plaza_metadata*

This data frame contains a list of species in PLAZA instances and associated
metadata, including taxonomic information.

```{r}
# Dicots ----
## PLAZA metadata
dicots_plaza <- read_tsv(
    "https://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_05/SpeciesInformation/species_information.csv.gz",
    show_col_types = FALSE, comment = "# "
) |>
    rename(PLAZA_ID = `#species`)

## Taxonomic information
dicots_tax <- taxize::classification(dicots_plaza$tax_id, db = "ncbi") |>
    format_classification()


# Monocots ----
## PLAZA metadata
monocots_plaza <- read_tsv(
    "https://ftp.psb.ugent.be/pub/plaza/plaza_public_monocots_05/SpeciesInformation/species_information.csv.gz",
    show_col_types = FALSE, comment = "# "
) |>
    rename(PLAZA_ID = `#species`)

## Taxonomic information
monocots_tax <- taxize::classification(monocots_plaza$tax_id, db = "ncbi") |>
    format_classification()


# Pico ----
## PLAZA metadata
pico_plaza <- read_delim(
    "ftp://ftp.psb.ugent.be/pub/plaza/plaza_pico_03/SpeciesInformation/species_information.csv.gz",
    delim = ";", show_col_types = FALSE, comment = "# "
) |>
    rename(PLAZA_ID = 1)

## Taxonomic information
pico_tax <- taxize::classification(pico_plaza$tax_id, db = "ncbi") |>
    format_classification()

# Diatoms ----
## PLAZA metadata
diatoms_plaza <- read_delim(
    "ftp://ftp.psb.ugent.be/pub/plaza/plaza_diatoms_01/SpeciesInformation/species_information.csv.gz",
    delim = ";", show_col_types = FALSE, comment = "# "
) |>
    rename(PLAZA_ID = 1)

## Taxonomic information
diatoms_tax <- taxize::classification(diatoms_plaza$tax_id, db = "ncbi") |>
    format_classification()

# Create final list of data frames
plaza_metadata <- list(
    Dicots = cbind(dicots_plaza, dicots_tax),
    Monocots = cbind(monocots_plaza, monocots_tax),
    Diatoms = cbind(diatoms_plaza, diatoms_tax),
    Pico = cbind(pico_plaza, pico_tax)
)

plaza_metadata <- lapply(plaza_metadata, function(x) {
    
    df <- x |>
        dplyr::select(
            PLAZA_ID, common_name, NCBI_name = ncbi_species, 
            NCBI_taxid = tax_id, family, order, class, source, 
            data_provider, pubmed_id
        )
    
    return(df)
})

# Save object
usethis::use_data(plaza_metadata, compress = "xz")
```

# Internal data

## sp2instance

This object contains a data frame of PLAZA species IDs and the PLAZA instance
where they can be found. If a species is found in multiple PLAZA 
instances (e.g., *Arabidopsis thaliana*), the following priority scheme
is used:

Dicots > Monocots > Diatoms > Pico.

The object `sp2instance` contains a data frame with 2 columns:
- `PLAZA_ID` (character): PLAZA species IDs as in `plaza_metadata`.
- `instance` (character): PLAZA instance. One of 'Dicots', 'Monocots', 
'Diatoms', or 'Pico'.

```{r}
data("plaza_metadata")
sp2instance <- plaza_metadata |>
    bind_rows(.id = "instance") |>
    select(PLAZA_ID, instance) |>
    distinct(PLAZA_ID, .keep_all = TRUE)

usethis::use_data(sp2instance, compress = "xz", internal = TRUE)
```

